As any work of such magnitude, a thesis cannot be completed by the efforts of
one individual alone. While there is room for only one name on the front cover,
this section allows the author to mention all the other people whose actions
contributed to the creation of this thesis.

First and foremost, I would like to thank my advisors, Hartwig Anzt and Enrique
Quintana-Ort\'{i} for taking on the challenge and the risk of guiding a fresh
graduate through the maze of academic research: your advice about every aspect
of this filed was invaluable; our frequent discussions were the most enjoyable
part of my job; and you selflessly invested more time than anyone could have
asked of you in smoothing my relocation and stay abroad.

This thesis would not be possible without the involvement of my instructors at
the University of Zagreb: Neven Krajina, Vedran Novakovi\'{c} and Sanja Singer,
who first introduced me to and provided detailed training in high performance
computing.

I am also grateful to my colleagues at Universidad Jaume I: Maria Barreda
Vay\`{a}, Rocio Carratal\'{a} S\'{a}ez, Adri\'{a}n Castell\'{o} Gimeno, Sandra
Catal\'{a}n Pallar\'{e}s, Sergio Iserte Agut and Andr\'{e}s Tom\'{a}s
Dom\'{i}nguez, who would often spend their valuable time as my advisors and
translators in dealings with local institutions. I especially want to thank
Rocio, for making me feel less of an outsider in a predominantly homogeneous
society.

I am deeply grateful to my parents and my family for their help in preserving my
emotional well-being during this adventure. I will always remember all the
family gatherings you organized in the rare occasions when I was able to come
home.

Finally, I thank my beloved Jelena. You have always been my moral and emotional
pillar, despite the strain of prolonged physical separation. For that, I am
eternally grateful.
